---
title: Improving Forecasts of Sockeye Salmon (*Oncorhynchus nerka*) with Modern Computational Tools.
subtitle: Supporting Information
author:
- Daniel Ovando*
- Curry Cunningham
- Peter Kuriyama
- Chris Boatright
- Ray Hilborn
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: default
  bookdown::word_document2:
    reference_docx: template.docx
  pdf_document: default
bibliography: ../references.bib
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
params:
  results_name: v1.1.0
linkcolor: blue
toc: FALSE
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}\linenumbers
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,  dev = "cairo_pdf", message = FALSE, warning = FALSE)
library(tidyverse)
library(patchwork)
library(cowplot)
library(magick)
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
library(here)
library(scales)
library(knitr)
library(kableExtra)
extrafont::loadfonts()

results_dir <- here("results", params$results_name)

pub_theme <-
  hrbrthemes::theme_ipsum(base_size = 10, axis_text_size = 8, axis_title_size = 12) +
  theme(
    panel.spacing = unit(0.5, "lines"),
    plot.margin = unit(rep(10, 4), units = "points")
  )

theme_set(pub_theme)


# load plots

load(file = file.path(results_dir, "plots.RData"))

load(file = file.path(results_dir, "performance.RData"))

load(file = file.path(results_dir, "forecasts.RData"))


```

```{r models}

benchmarks <- system_performance %>% 
  filter(model == "fri" | model == "lag") %>% 
  select(model, system, rmse) %>% 
  pivot_wider(names_from = model, values_from = "rmse")

age_benchmarks <- age_performance %>% 
  filter(model == "fri" | model == "lag") %>% 
  select(model, age_group, rmse) %>% 
  pivot_wider(names_from = model, values_from = "rmse")


sys_performers <- system_performance %>% 
  filter(!model %in% c("fri"), !str_detect(model,"ensemble")) %>% 
  group_by(system) %>% 
  filter(srmse == min(srmse))

comp_sys_improvement <- sys_performers %>% 
  select(model, system, rmse) %>% 
  left_join(benchmarks, by = "system") %>% 
  ungroup() %>% 
  summarise(imp_on_fri = -mean(rmse / fri - 1),
            min_imp_on_fri = min(-(rmse / fri - 1)),
            max_imp_on_fri = max(-(rmse / fri - 1)),
            imp_on_lag =  -mean(rmse / lag - 1))

tmp <- sys_performers %>% 
  select(model, system, rmse) %>% 
  left_join(benchmarks, by = "system") %>% 
  ungroup() %>% 
  mutate(imp_on_fri = -(rmse / fri - 1),
            min_imp_on_fri = (-(rmse / fri - 1)),
            max_imp_on_fri = (-(rmse / fri - 1)),
            imp_on_lag =  -(rmse / lag - 1))

age_performers <- age_performance %>% 
  filter(!model %in% c("fri", "boost_tree_ensemble")) %>% 
  group_by(age_group) %>% 
  filter(srmse == min(srmse))

tmp2 <- age_performers %>% 
  select(model, age_group, rmse) %>% 
  left_join(age_benchmarks, by = "age_group") %>% 
  ungroup() %>% 
  mutate(imp_on_fri = -(rmse / fri - 1),
            min_imp_on_fri = (-(rmse / fri - 1)),
            max_imp_on_fri = (-(rmse / fri - 1)),
            imp_on_lag =  -(rmse / lag - 1))

a = tmp %>%
  ggplot(aes(reorder(system, -imp_on_fri), imp_on_fri, fill = model)) +
  geom_hline(aes(yintercept = 0)) +
  geom_col(color = "black") +
  scale_y_percent(name = "% Improvement on Traditional FRI Methods") +
  scale_x_discrete(name = '')+
  fishualize::scale_fill_fish_d(name = 'Best Model', option = "Trimma_lantana") + 
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 14)) 

ggsave( "rep_fig_1.pdf",a, device = cairo_pdf, width = 6, height = 4)

b = tmp2 %>%
  mutate(age_group = str_replace_all(age_group,"_",".")) %>% 
  ggplot(aes(reorder(age_group, -imp_on_fri), imp_on_fri, fill = model)) +
  geom_hline(aes(yintercept = 0)) +
  geom_col(color = "black") +
  scale_y_percent(name = "% Improvement on Traditional FRI Methods") +
  scale_x_discrete(name = '')+
  fishualize::scale_fill_fish_d(name = 'Best Model', option = "Trimma_lantana") + 
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 14)) 

ggsave( "rep_fig_2.pdf",b, device = cairo_pdf, width = 6, height = 4)



d <- total_forecast %>%
  filter(model %in% c("boost_tree", "fri")) %>%
  group_by(year) %>%
  mutate(observed = mean(observed)) %>%
  pivot_wider(names_from = "model", values_from = "forecast") %>%
  ggplot() +
  geom_hline(aes(yintercept = 0)) +
  geom_area(aes(year, observed),fill = "lightgrey") +
  geom_point(aes(year, boost_tree, fill = "boost_tree"), shape = 21, size = 4,alpha = 0.9) +
  geom_point(aes(year, fri, fill = "UW-FRI"), shape = 21, size = 4,alpha =0.9) + 
  fishualize::scale_fill_fish_d(name = '', option = "Trimma_lantana") + 
  scale_y_continuous(name = "Total Returns") + 
  scale_x_continuous(name = 'Year') + 
    theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 14)) 


ggsave( "rep_fig_3.pdf",d, device = cairo_pdf, width = 6, height = 4)

fri_system_mape <- system_forecast %>% 
  filter(model == "fri") %>%
  group_by(model) %>% 
  summarise(mape = yardstick::mape_vec(observed, forecast))

fri_age_mape <- age_forecast %>% 
  filter(model == "fri") %>%
  group_by(model, age_group) %>% 
  summarise(mape = yardstick::mape_vec(observed, forecast))

# system_forecast %>% 
#   ggplot(aes(observed, forecast)) + 
#   geom_point() + 
#   geom_abline(aes(slope = 1, intercept = 0)) +
#   facet_wrap(~model)


mean_srmse_delta <- system_performance %>% 
    filter(!model %in% c("fri", "boost_tree_ensemble")) %>% 
  group_by(system) %>% 
  mutate(rmse_rank = rank(rmse)) %>% 
  filter(rmse_rank <= 2) %>% 
  arrange(system) %>% 
  group_by(system) %>% 
  summarise(delta  = srmse[rmse_rank == 2] - srmse[rmse_rank == 1]) %>% 
  ungroup() %>% 
  summarise(mmd = mean(delta))

comp_sys_improvement <- sys_performers %>% 
  select(model, system, rmse) %>% 
  left_join(benchmarks, by = "system") %>% 
  ungroup() %>% 
  summarise(imp_on_fri = -mean(rmse / fri - 1),
            min_imp_on_fri = min(-(rmse / fri - 1)),
            max_imp_on_fri = max(-(rmse / fri - 1)),
            imp_on_lag =  -mean(rmse / lag - 1))


sys_improv <- system_performance %>% 
  select(model, system, srmse) %>% 
  pivot_wider(names_from = model, values_from = srmse) %>% 
  pivot_longer(-c(system, fri), names_to = "model", values_to = "srmse") %>% 
  group_by(system) %>% 
  filter(srmse == min(srmse)) %>% 
  ungroup() %>% 
  summarise(srmse_improv = mean((srmse - fri) / fri))

age_improv <- age_performance %>% 
  select(model, age_group, srmse) %>% 
  pivot_wider(names_from = model, values_from = srmse) %>% 
  pivot_longer(-c(age_group, fri), names_to = "model", values_to = "srmse") %>% 
  group_by(age_group) %>% 
  filter(srmse == min(srmse)) %>% 
  ungroup() %>% 
  summarise(srmse_improv = mean((srmse - fri) / fri))


```


\renewcommand{\thefigure}{S\arabic{figure}}

\setcounter{figure}{0}

\renewcommand{\thetable}{S\arabic{table}}

\setcounter{table}{0}

```{r srmse-fig, fig.cap="Scaled root mean squared error (SRMSE) for each model type (rows) by age group and system (columns).", fig.height=3}
srmse_summary_figure + 
  theme(axis.text.x = element_text(size = 6))
```

```{r srmse-fig-two, fig.cap="Best performing model in terms of scaled root mean squared error (SRMSE) for each age group and river system. Color indicates the percent improvement in SRMSE between the best performing model and the historic published Fisheries Research Institute (FRI) forecasts."}
age_system_performance_figure
```

+-----------------------------------+-----------------------------------------------------------------------------------------------+-------------------------------------------------------+
| Data Type                         | Description                                                                                   | Dataset ID                                            |
+===================================+===============================================================================================+=======================================================+
| Sea Surface Temperature           | HadISST Average Sea Surface Temperature, 1°, Global, Monthly, 1870-present                    | erdHadISST                                            |
+-----------------------------------+-----------------------------------------------------------------------------------------------+-------------------------------------------------------+
| Sea Level Pressure                | Sea Level Pressure Monthly Mean at Surface, ICOADS, 1-degree, Enhanced, Monthly, 1960-present | esrlIcoads1ge                                         |
+-----------------------------------+-----------------------------------------------------------------------------------------------+-------------------------------------------------------+
| Upwelling                         | u-wind Stress Monthly Mean at Surface, ICOADS, 1-degree, Enhanced, Monthly, 1960-present      | esrlIcoads1ge                                         |
+-----------------------------------+-----------------------------------------------------------------------------------------------+-------------------------------------------------------+
|                                   |                                                                                               |                                                       |
+-----------------------------------+-----------------------------------------------------------------------------------------------+-------------------------------------------------------+
| Pacific Decadal Oscillation Index | PDO Index                                                                                     | <http://research.jisao.washington.edu/pdo/PDO.latest> |
+-----------------------------------+-----------------------------------------------------------------------------------------------+-------------------------------------------------------+

: (\#tab:env-data) Environmental Covariates Evaluated

+----------------------------------------+-------------------------------------------------------------------------------+
| Model Name                             | Description                                                                   |
+========================================+===============================================================================+
| Boosted Regression Trees (boost\_tree) | An ensemble of boosted regression trees                                       |
+----------------------------------------+-------------------------------------------------------------------------------+
| Random Forest (rand\_forest)           | An ensemble of regression trees                                               |
+----------------------------------------+-------------------------------------------------------------------------------+
| Recurrent Neural Network (rnn)         | A neural network with a memory component implemented through recurrent layers |
+----------------------------------------+-------------------------------------------------------------------------------+
| Dynamic Linear Model (dlm)             | Sibling regression with dynamic components                                    |
+----------------------------------------+-------------------------------------------------------------------------------+
| Empirical Dynamic Modeling  (multiview)| Simplex projection based on lagged values                                     |
+----------------------------------------+-------------------------------------------------------------------------------+
| Lag-1                       | Forecast next year equal to observed returns last year                        |
+----------------------------------------+-------------------------------------------------------------------------------+
| Ensemble                               | A random forest ensemble made of other candidate models             |
+----------------------------------------+-------------------------------------------------------------------------------+
|                                        |                                                                               |
+----------------------------------------+-------------------------------------------------------------------------------+

: (\#tab:models) Candidate forecasting models evaluated



+----------------------------------------+-------------------------------------------------------------------------------+
| Model Name                             | Description                                                                   |
+========================================+===============================================================================+
| Boosted Regression Trees (boost\_tree) | An ensemble of boosted regression trees                                       |
+----------------------------------------+-------------------------------------------------------------------------------+
| Random Forest (rand\_forest)           | An ensemble of regression trees                                               |
+----------------------------------------+-------------------------------------------------------------------------------+
| Recurrent Neural Network (rnn)         | A neural network with a memory component implemented through recurrent layers |
+----------------------------------------+-------------------------------------------------------------------------------+
| Dynamic Linear Model (dlm)             | Sibling regression with dynamic components                                    |
+----------------------------------------+-------------------------------------------------------------------------------+
| Empirical Dynamic Modeling  (multiview)| Simplex projection based on lagged values                                     |
+----------------------------------------+-------------------------------------------------------------------------------+
| Lag-1                       | Forecast next year equal to observed returns last year                        |
+----------------------------------------+-------------------------------------------------------------------------------+
| Ensemble                               | A random forest ensemble made of other candidate models             |
+----------------------------------------+-------------------------------------------------------------------------------+
|                                        |                                                                               |
+----------------------------------------+-------------------------------------------------------------------------------+

: (\#tab:boost-tree-params) Range of tuning parameters tested for boosted regression tree



\newpage


```{r tuning-table, results="asis"}


boost_tree_grid <-
  tribble(
    ~ "Variable",
    ~ "Abbreviation",
    ~ "Range Tested",
    "Minimum node size",
    "min_n",
    "[2,10]",
    "Maximum tree depth",
    "tree_depth",
    "[2,15]",
    "Learning rate",
    "learn_rate",
    "[0.05,0.6]",
    "Number of randomly sampled predictors",
    "mtry",
    "[1,# of Predictors]",
    "Loss reduction for further splitting",
    "loss_reduction",
    "[1e-15,2]",
    "Number of trees",
    "trees",
    "[500,2000]"
  ) %>% 
  mutate(model = "Boosted Regression Tree (xgboost)")



rf_grid <-
  tribble(
    ~ "Variable",
    ~ "Abbreviation",
    ~ "Range Tested",
    "Minimum node size",
    "min_n",
    "[2,10]",
    "Number of randomly sampled predictors",
    "mtry",
    "[1,# of Predictors]",
    "Number of trees",
    "trees",
    "[500,2000]"
  ) %>% 
  mutate(model = "Random forest (ranger)")

grid_summary <- boost_tree_grid %>% 
  bind_rows(rf_grid)


kbl(grid_summary, caption = "Ranges of tuning parameters tested for machine learning models", booktabs = TRUE)



```

