---
output: github_document
params: 
  results_name: ["v0.5.4"]
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)


```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,  dev = "cairo_pdf", message = FALSE, warning = FALSE)
library(tidyverse)
library(patchwork)
library(cowplot)
library(magick)
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
library(here)
library(scales)
extrafont::loadfonts()

results_dir <- here("results", params$results_name)

pub_theme <-
  hrbrthemes::theme_ipsum(base_size = 10, axis_text_size = 8, axis_title_size = 12) +
  theme(
    panel.spacing = unit(0.5, "lines"),
    plot.margin = unit(rep(10, 4), units = "points")
  )

theme_set(pub_theme)


# load plots

load(file = file.path(results_dir, "plots.RData"))

load(file = file.path(results_dir, "performance.RData"))

load(file = file.path(results_dir, "forecasts.RData"))


```

# Salmon Forecast Paper

<!-- badges: start -->
<!-- badges: end -->

Materials needed to produce results for the loosely titled "Evaluating the Potential of Computer-Age Statistical Methods in Ecological Forecasting"


## Reproducing Results

All materials needed to reproduce our results and manuscript are contained in this repository. In order to reproduce

1. Fork the repository and clone to your machine

2. Open R and set your working directory of the cloned repository (or just use RStudio projects)

3. This project is set up with [`renv`](https://rstudio.github.io/renv/articles/renv.html) to manage package dependencies. Inside R (and with your working directory set correctly) run `renv::restore()`. This will install the correct versions of all the packages needed to replicate our results. Packages are installed in a stand-alone project library for this paper, and will not affect your installed R packages anywhere else. 

4. Open `make-salmon-forecast-paper.R` and
  - set the name of the run in the call to `prep_run`
  - ensure that 

```{r, eval = FALSE, echo = TRUE}
run_edm_forecast <- TRUE

run_dlm_forecast <- TRUE

run_ml_forecast <- TRUE

fit_statistical_ensemble <- TRUE

run_importance <- TRUE

knit_manuscript <- TRUE

```

5. source `make-salmon-forecast-paper.R`

6. The resulting manuscript will be compiled in the documents folder

## Abstract

  Salmon are an ecological and economic keystone of Alaska. Effective management of these species depends on accurate pre-season forecasts, allowing fishery regulations to be set appropriately and fishing operations to plan their economic activities accordingly.  These predictions are challenging to make though, given the complex set of dynamic factors that affect numbers of returning salmon, from egg production to predation to environmental drivers. Computer-age statistical methods such as dynamic linear modeling, machine learning and empirical dynamical modeling have fundamentally transformed many aspects of our society by providing previously impossible levels of predictive accuracy to tasks from digital advertising to weather forecasting. However, these methods underutilized as applied tools for predicting and managing natural resources. We utilize data from the salmon populations of Bristol Bay,  Alaska to evaluate the potential for computer-age statistical methods to improve the accuracy of predictions of annual returns. Salmon present an ideal case study due to the presence of long and accurate time series of annual returns, and the ecological and economical  importance of reliable predictions to commercial fisheries and subsistence-dependent communities. We demonstrate how computer-age methods can help researchers and managers provide improvements in forecast skill, and identify when the bottleneck predictive skill is model design versus data. We also show that these methods may not be a panacea for improving predictive performance when confronted with time series of short duration, as are often found in analysis of population dynamics and ecological forecasting applications at annual time steps for aquatic and terrestrial species.  


## Sample Results

```{r}
system_forecast_figure
```


```{r}
age_system_performance_figure
```


```{r}
total_forecast_figure
```

